---
title: "MvLA Report 2020"
author: "Michael Haskins"
date: "28/11/2020"
output: 
 ## pdf_document:
  ##number_sections: true
 ##   df_print: kable

## word_document

 rmdformats::readthedown:
    lightbox: true
    highlight: kate

##fontsize: 11pt  
##geometry: margin=1in  
---

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
#knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(rmdformats)
library(tinytex)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```




# Introduction

The heart data set (file [heart_data.csv](heart_data.csv) ) consists of nine variables, each measuring different physiological characteristics of patients at risk of heart
disease. Five variables are continuous, and three are binary. One variable,
Slope, is ordinal, but will be treated as continuous. The tenth
variable in the data set is Class, and this denotes whether heart disease is
absent (1) or present (2). 

This report will consist of two sections;

   1. Cluster Analysis:
      1. Identifying subsets present in the data
      2. Identifying subsets at higher risk of heart disease
   2. Classification:
      1. Predicting the patients with heart disease
    



```{r message=FALSE, echo=FALSE}
##May need to require certain packages here
require("knitr") 
heart = read.csv("heart_data.csv")

```

# Cluster Analysis

## Identifying subsets present in the data

Section Aim:

  * To identify subsets of patients in the data set who have similar physiological characteristics. and determine if these subsets are at higher risk of heart disease.


## Identifying subsets at higher risk of heart disease

Section Aim: 

  * To determine if any of the identified subsets are at higher risk of heart disease.

```{r message=FALSE, echo=FALSE}
## echo=FALSE hides the code in the pdf or html


```

# Classification

## Predicting the patients with heart disease

  Section Aim: 

  * To provide accurate predictions of patients with heart disease.
