---
title: "STU33010: Mid Term Assignment 2020"
author: "Michael Haskins"
date: "14/11/2020"
output: 
  rmdformats::readthedown:
    lightbox: true
    highlight: kate
  
   
---

```{r setup, include=FALSE,warning=FALSE, message=FALSE}
#knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(rmdformats)

## Global options
options(max.print="75")
opts_chunk$set(echo=TRUE,
	             cache=FALSE,
               prompt=FALSE,
               tidy=TRUE,
               comment=NA,
               message=FALSE,
               warning=FALSE)
opts_knit$set(width=75)
```




# Time series data

The   time series dataset (file [pop.csv](pop.csv) ) corresponds to several time series of population size for the UK till year 2018.


## Reading the time series from a *.csv file



```{r message=FALSE}
# make sure to type in Rstudio console: install.packages("fma")
require("fma")
require("knitr") # install package if not installed
myts<-read.csv("pop.csv")

kable(myts)

```



## Example for visualisation 


```{r message=FALSE }

mytsScotland<-ts(myts$Scotland.population, start="1971",end="2018",frequency=1)

tsdisplay(mytsScotland)
```



# Holt-Winters algorithms (20 Marks)

**Find the best Holt-Winters algorithms to predict the populations of England, Scotland, Northern Ireland and Wales (4 time series) for the years  2019 and 2020.**

* Write a brief report with R code and  comments  
  + explaining  the methodology you have used to select the best algorithm for each time series (10 Marks).
  + Compute the forecasts (2019 and 2020) for each of the 4 time series (5 Marks)
  + With a maximum of 100 words, what warnings would you write to comment on  the accuracy of these forecasts (5 Marks). 
  
* Submit your report (html+Rmd, or pdf+R files)  in blackboard and also fill in (copy-paste) your answers in this [online form](https://forms.office.com/Pages/ResponsePage.aspx?id=jb6V1Qaz9EWAZJ5bgvvlK9aLUzeim5BDngtcfukjGHJUOFFTSjdTRTJWWlVFRTM0SlhFODZEN0U5Uy4u). 



# Submission Format

For all students in the module with an email address of the form StudentName@tcd.ie


To be submitted on **Blackboard**: 

- 2 files named *StudentName.Rmd* and *StudentName.html*   

or (if Rmd does not work)

- 2 files named *StudentName.R* (code)  and *StudentName.pdf* (report)

An additional [online form](https://forms.office.com/Pages/ResponsePage.aspx?id=jb6V1Qaz9EWAZJ5bgvvlK9aLUzeim5BDngtcfukjGHJUOFFTSjdTRTJWWlVFRTM0SlhFODZEN0U5Uy4u) needs to be filled in as well with your answers.


# Marking 

The submission is marked :  

- for correctness, 

- quality of expression (e.g. R, English) and explanations provided.

- brevity and clarity are  valued.    

- and following the instructions.



